---
title: "blusteR: Clustering BCR Sequences by Shared Antigen Specificity"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{blusteR: Clustering BCR Sequences by Shared Antigen Specificity}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

## Introduction

blusteR (**B**-cell **L**ymphocyte **I**nteraction by **P**aratope **H**otspots)
clusters B-cell receptor sequences that likely recognise the same antigen.  It
adapts the GLIPH2 algorithm — originally designed for T-cell receptors — to the
unique biology of antibodies.

### Why not just use GLIPH2 on BCR data?

BCR and TCR biology differ in several fundamental ways that require a
purpose-built method:

1. **Somatic hypermutation (SHM).**
   After initial antigen encounter, B cells undergo iterative rounds of
   mutation in their variable regions.  Two B cells that started from the
   same germline rearrangement may differ by dozens of point mutations yet
   still bind the same epitope.  GLIPH2's Hamming-distance metric treats
   all substitutions equally, penalising conservative SHM events (e.g.
   serine → threonine) just as heavily as radical changes.  blusteR uses
   BLOSUM62-weighted distances that distinguish SHM-typical conservative
   substitutions from specificity-altering ones.

2. **Longer CDR3 regions.**
   Human CDR3H averages ~15 amino acids (vs ~13 for CDR3β in TCRs) with
   greater length variance.  GLIPH2's short k-mers (2–3 aa) are
   underpowered for longer sequences; blusteR uses 4–5 aa k-mers and
   tolerates ±1 aa length differences in global comparisons.

3. **Paired chain information.**
   10X Genomics single-cell VDJ provides matched heavy + light chain
   sequences.  While GLIPH2 focuses on CDR3β with optional CDR3α, blusteR
   natively supports paired heavy + light chain analysis with weighted
   distance scoring (CDR3H contributes 70%, CDR3L 30%).

4. **Clonal expansion.**
   B-cell clonal families can dominate a sample.  blusteR collapses clonal
   relatives before clustering to prevent clonal expansion from inflating
   cluster membership.

5. **Different reference databases.**
   TCR specificity databases (VDJdb, McPAS-TCR) are irrelevant for BCR
   analysis.  blusteR uses the Observed Antibody Space (OAS) for background
   statistics, IEDB for B-cell epitope annotation, and SAbDab for
   structurally validated antibody-antigen pairs.

## Installation

```{r install}
# Bioconductor dependency
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("Biostrings")

# CRAN dependencies
install.packages(c("data.table", "stringdist", "igraph",
                    "ggplot2", "jsonlite", "httr"))

# Install blusteR
install.packages("/path/to/blusteR", repos = NULL, type = "source")
```

## Loading data

### From 10X Genomics Cell Ranger VDJ

```{r load_10x}
library(blusteR)

# Point to the Cell Ranger VDJ output directory
bcr <- load_10x_bcr(
  path = "/data/cellranger_vdj/sample1/outs/",
  sample_id = "Patient1_baseline"
)

head(bcr)
#>                     cell_id         sample_id chain v_gene  ... cdr3_aa
#> 1: Patient1_baseline_AAACCTGAGCTAGCCC-1 Patient1_baseline   IGH IGHV3-30 ... CARDGYSSGWYFDY
#> 2: Patient1_baseline_AAACCTGAGCTAGCCC-1 Patient1_baseline   IGK IGKV1-39 ... CQQSYSTPLTF
```

### From scRepertoire

```{r load_scr}
library(scRepertoire)

contig_list <- list(
  read.csv("sample1_filtered_contig_annotations.csv"),
  read.csv("sample2_filtered_contig_annotations.csv")
)

combined <- combineBCR(
  contig_list,
  samples = c("Donor1", "Donor2")
)

# Convert to blusteR format
bcr <- load_screpertoire(combined, sample_ids = c("Donor1", "Donor2"))
```

### Multi-sample design

For discovering convergent antibody responses across individuals:

```{r multi}
samples <- list(
  list(path = "donor1_pre/",  id = "D1_pre"),
  list(path = "donor1_post/", id = "D1_post"),
  list(path = "donor2_pre/",  id = "D2_pre"),
  list(path = "donor2_post/", id = "D2_post")
)

bcr_all <- data.table::rbindlist(lapply(samples, function(s) {
  load_10x_bcr(s$path, sample_id = s$id)
}))
```

## Running the pipeline

### One-call analysis

```{r pipeline}
result <- bluster(
  bcr_data          = bcr_all,
  chain             = "heavy",       # or "paired" for H+L
  global_scoring    = "blosum",      # SHM-aware distances
  min_cluster_size  = 3,
  collapse_clones   = TRUE,          # avoid clonal inflation
  annotate          = TRUE,          # match against IEDB/SAbDab
  species           = "human"
)
```

### Step-by-step control

```{r steps}
# 1. Build reference
ref <- build_reference(species = "human")

# 2. Filter
bcr_filt <- filter_bcr(bcr_all, min_cdr3_len = 5, max_cdr3_len = 30)

# 3. Motif discovery
motifs <- bluster_motifs(bcr_filt, reference = ref, chain = "heavy",
                       k_sizes = c(4, 5), min_freq = 3)

# 4. Global similarity
global <- bluster_global(bcr_filt, max_dist = 2, chain = "heavy",
                        scoring = "blosum")

# 5. Cluster
result <- bluster_cluster(bcr_filt, motifs, global,
                         min_cluster_size = 3,
                         collapse_clones = TRUE)

# 6. Annotate
result <- annotate_epitopes(result, reference = ref, bcr_data = bcr_filt)
```

## Interpreting results

### Cluster summaries

```{r summary}
summarize_clusters(result)

# Access the cluster table directly
result$clusters[order(-bluster_score)][1:5]
```

Key columns in `$clusters`:

- **n_members**: Cells in the cluster (after clone collapsing)
- **n_clones**: Distinct clonotypes (high = convergent selection)
- **n_samples**: Donors/timepoints represented
- **cdr3h_consensus**: Representative CDR3H sequence
- **top_vh**: Most common IGHV gene
- **bluster_score**: Composite confidence score (0–1)

### Epitope annotations

```{r annotations}
result$annotations
#>    cluster_id     query_cdr3      match_cdr3 match_type distance    antigen
#> 1:   blusteR_14 CARDGYSSGWYFDY CARDGYSSGWYFDY      exact        0 SARS-CoV-2 Spike
```

### Cluster membership

```{r membership}
# Which cells are in each cluster?
result$membership[cluster_id == "blusteR_1"]

# Merge back with original metadata
merged <- merge(result$membership, bcr_all, by = "cell_id")
```

## Visualisation

```{r plots, fig.width=8, fig.height=6}
# Network view of all specificity groups
plot_cluster_network(result)

# Enriched CDR3 motifs
plot_motif_logo(result)

# V-gene usage across clusters
plot_vgene_usage(result)

# Four-panel overview
plot_cluster_summary(result)
```

## Exporting results

```{r export}
export_clusters(result, prefix = "my_study", dir = "bluster_output/")
# Creates:
#   bluster_output/my_study_clusters.csv
#   bluster_output/my_study_membership.csv
#   bluster_output/my_study_motifs.csv
#   bluster_output/my_study_annotations.csv
```

## Parameter guidance

| Parameter | Default | When to change |
|-----------|---------|----------------|
| `chain` | "heavy" | Use "paired" if you have matched H+L chains |
| `k_sizes` | c(4,5) | Try c(3,4) for shorter CDR3s (e.g. mouse IgM) |
| `global_max_dist` | 2 | Increase to 3 for more permissive clustering |
| `global_scoring` | "blosum" | Use "hamming" for speed on large datasets |
| `min_cluster_size` | 3 | Increase for stringent analysis (e.g. 5) |
| `clustering_method` | "components" | Use "louvain" for >5000 unique CDR3s |
| `collapse_clones` | TRUE | Set FALSE only if clonal info is unreliable |

## Session info

```{r sessioninfo}
sessionInfo()
```
